\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-13 analysis/manuscript.tex
continue introduction

For over two decades, hybrid open access journal publishing, which makes
some articles freely available under an open license while others remain
behind a paywall, has been discussed as a model for transitioning the
subscription system to full open access (Prosser, 2003). The idea was
that when journals publish more and more open access articles, they
could reduce revenues from subscriptions, while libraries and funders
could change their funding models and shift expenditures from
subscription to open access. However, initial approaches, mainly based
on publication fees, also called article processing charges (APCs), did
not contribute substantially to an large open access uptake. In 2012 ddd
- at the same time

Tthe introduction of transformative agreements, a collective term for
agreements starting from mid-2015, has resulted in a new push for hybrid
open access. In particular, the Max Planck Society's Berlin 2020
initiative and plan has advocated these types of agreements as a
transitional means to progress with open access. As a result, more than
800 transformative agreements have been tracked so far, which have been
instrumental in funding activities and controversies. today, all
publsiehrs offer such agreements

Advocates of these agreements argue that they not only overcome the
issue of double-dipping but also help to shift library spending from
subscriptions to publications, streamline expenditures, and offer more
value than big deals, which focus on large publishers. A few agreements
can impact an institution's open access share, resulting in a
significant proportion of open access content. increase transparency

However, opponents of transformative agreements perceive continued
reliance on big deals as problematic. They argue that the focus on large
publishers perpetuates the business relationships that led to the serial
crisis and lacks evidence that this leads to a reduction of In addition,
concerns over inequality abound. could not benn mass removal

@@

\begin{center}\includegraphics[width=0.99\linewidth,]{fig/countrypatch-1} \end{center}

Figure \ref{fig:countrypatch}A shows the development of hybrid open
access publishing by countries, comparing the OECD area with the BRICS,
an intergovernmental organisation, which comprised the countries Brazil,
Russia, India, China and South Africa in 2022. The residual category
``Other'' includes the remaining countries. From 2018 to 2022, the
proportion of open access in hybrid journals increased from 6.1\% in
2018 to 26\% in 2022. On the other hand, BRICS recorded an low uptake,
from 1.6\% in 2018 to 3.7\% in 2022.\\
Despite rise of open access across OECD countries, the overall
publication output decreased sharply, dropping to 786,903 in 2022 after
peaking 892,197 articles in 2020. In stark contrast, the number of
articles published in hybrid journals by first authors affiliated with
institutions from BRICS countries increased steadily over the years,
more than doubling from 356,632 in 2018 to 786,903 in 2022. Upon closer
examination, this trend can be observed across all big three publishers,
although the shift towards BRICS is particularly evident in Elsevier's
hybrid journal portfolio, in particular with regard to articles
published in Physical Sciences journals (see Supplement). While OECD
publication output in Elsevier's Physical Sciences journals declined
from 112,822 articles in 2018 to 103,766 in 2022, BRICS output increased
from 104,654 to 171,713 in the same five-year period. Furthermore, OECD
publication output in Health Science Journals and Life Science journals
stagnated across the investigated hybrid journal portfolios after a peak
in 2020.

To illustrate the situation in 2022, \ref{fig:countrypatch}B compares
total publication output with the number of open access articles. With
391,530 articles, China was the most productive country, followed by the
United States (268,965 articles) and India (87,428 articles). In
contrast, West and Nord European countries published a considerable high
number of open access articles, mainly due to transformative agreements.
Particularly, Germany, Great Britain, the Netherlands, Sweden,
Switzerland and Spain recorded an above-average open access share as
indicated by the linear trend line. As represented by the point size, as
well as it can been seen in Figure \ref{fig:country_patch}C,
transformative agreements contributed to this market position of these
countries. Interestingly, the United States had a notable open access
market share of 15\%, although transformative agreements contributed to
a lesser extent. Similarly, China's open access market share of 7.2 in
2022 was comparable to that of the Netherlands, which was (7.1\%,).

\begin{center}\includegraphics[width=0.99\linewidth,]{fig/countrytop20plot-1} \end{center}

Figure \ref{fig:countrytop20plot} illustrates the development of hybrid
open access from 2018 to 2022, highlighting the top 20 most productive
countries in terms of articles published in hybrid journals that were
included in transformative agreements over the five-year period.
Notably, The Netherlands (27\%), Sweden (24\%), Poland (17\%) and Great
Britain (17\%)) exhibited a relatively high level of uptake in 2018
which continued to increase in the following years. In 2022, Sweden had
the highest proportion of open-access articles relative to its
publication output (78\%), followed by the Netherlands (67\%) and
Switzerland (57\%), with these countries benefiting from transformative
agreements. In Germany, however, hybrid open access only began to
increase from 2019 onwards after the successful negotiation of
nationwide agreements with Wiley (July 2019) and Springer Nature
(January 2020). Prior to this, only a few organisations had agreements
in place, for the example the Max Planck Society with Springer Compact.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Borrego_2020}{}}%
Borrego, Á., Anglada, L., \& Abadal, E. (2020). Transformative
agreements: Do they pave the way to open access? \emph{Learned
Publishing}. \url{https://doi.org/10.1002/leap.1347}

@@

@@

@@

@@

@@

@@

@@

>>>>

For over two decades, hybrid open access journal publishing, which makes
some articles openly available while others remain behind a paywall, has
been discussed as a model for transitioning the subscription system to
full open access (Prosser, 2003). The idea was that when journals
publish more and more open access articles, they could reduce revenues
from subscriptions, while libraries and funders could change their
funding models and shift expenditures from subscription to open access.
However, initial approaches, mainly based on publication fees, also
called article processing charges (APCs), did not contribute
substantially to an large open access uptake. In 2009, Springer reported
open access to 1\% of articles in hybrid journals (Dallmeier-Tiessen et
al., 2010). Other publishers also recorded a low uptake. In 2011, only
1-2\% of articles in over 4,300 hybrid journals from large publishers
were open access (Björk, 2012), growing to around 4\% between 2011 and
2013 (Laakso \& Björk, 2016).

With the introduction of central funding mechanisms for hybrid open
access publication fees in a few countries since 2012, notably in the
UK, an increase in open access could be observed (Björk, 2017; Jubb et
al., 2017; Robinson-Garcia et al., 2020). However, this shift in funding
policy towards hybrid open access also added to the overall cost of
publishing, which includes subscriptions spending and the administrative
efforts required to handle payments (Pinfield et al., 2016). Moreover,
established large commercial publishers, which already dominated the
publishing market (Larivière et al., 2015), disproportionately benefited
from hybrid open access funding in comparison to full open access
publishers (Butler et al., 2023; Jahn \& Tullney, 2016; Shu \&
Larivière, 2023).

As a consequence, libraries and their consortia began to develop
licensing strategies aimed at avoiding such `double dipping' scenarios,
in which well-established commercial publishers gain twice from reading
and publication fees, as well as to increase publisher-provided
immediate open access (BJÖRK \& SOLOMON, 2014; Schimmer et al., 2015).
These considerations resulted into transformative agreements, which
cover a broad range of successfully negotiated contracts between library
consortia and publishers from mid-2010s onwards where institutional
spending for subscriptions are re-purposed for open access publishing
(Borrego et al., 2020). These transformative agreements seek to control
costs while allowing a transitional phase for publishing more open
access articles (Hinchliffe, 2019). At the same time, they aim at a
higher degree of transparency than previous subscription-only Big Deals,
which bundled access to large journal portfolios of commercial
publishers (Bergstrom et al., 2014). However, the focus of
transformative agreements remains primarily on these publishers with
large hybrid journals and subscription-only portfolios (Borrego et al.,
2020). For example, the nationwide DEAL consortium in Germany, which
includes universities and academic institutions, collectively cancelled
their subscriptions with Elsevier from 2017 onwards to negotiate a
transformative agreement (Fraser et al., 2023).

The introduction of transformative agreements aligns with funding policy
changes, such as the decision made by the cOAlition S, a consortium of
national funders including the European Commission, to no longer provide
financial support for individual publication fees when publishing in
hybrid journals. According ot its Plan S, they only accept hybrid open
access through transformative agreements ``during a transition period
that should be as short as possible'' (Schiltz, 2018). Specifically,
cOAlition S funders agreed to support hybrid open access solely through
transformative agreements from 1 January 2021, until the end of 2024.
The German Research Foundation (DFG) has also extended its financial
support for hybrid open access through transformative agreements,
despite not being part of cOAlition S. Previously, the DFG only provided
funding for full open access journals (Jahn \& Tullney, 2016).

Advocates of these agreements argue that they not only overcome the
issue standardiazation of double-dipping but also help to shift library
spending from subscriptions to publications, streamline expenditures,
and offer more value than big deals, which focus on large publishers. A
few agreements can impact an institution's open access share, resulting
in a significant proportion of open access content. increase
transparency

However, opponents of transformative agreements perceive continued
reliance on big deals as problematic. They argue that the focus on large
publishers perpetuates the business relationships that led to the serial
crisis and lacks evidence that this leads to a reduction of In addition,
concerns over inequality abound. trnasparency differ and now nsns

could not benn mass removal

@@

{\centering \includegraphics[width=0.99\linewidth,]{fig/country_patch-1}

\caption{Development of hybrid open access publishing by countries}\label{fig:country_patch}
\end{figure}

Figure \ref{fig:country_patch}A shows the development of hybrid open
access publishing by countries, comparing the OECD area with the BRICS,
an intergovernmental organisation, which comprised the countries Brazil,
Russia, India, China and South Africa in 2022. The residual category
``Other'' includes the remaining countries. From 2018 to 2022, the
proportion of open access in hybrid journals increased from 6.1\% in
2018 to 26\% in 2022. On the other hand, BRICS recorded an low uptake,
from 1.6\% in 2018 to 3.7\% in 2022.\\
Despite rise of open access across OECD countries, the overall
publication output decreased sharply, dropping to 786,903 in 2022 after
peaking 892,197 articles in 2020. In stark contrast, the number of
articles published in hybrid journals by first authors affiliated with
institutions from BRICS countries increased steadily over the years,
more than doubling from 356,632 in 2018 to 786,903 in 2022. Upon closer
examination, this trend can be observed across all big three publishers,
although the shift towards BRICS is particularly evident in Elsevier's
hybrid journal portfolio, in particular with regard to articles
published in Physical Sciences journals (see Supplement). While OECD
publication output in Elsevier's Physical Sciences journals declined
from 112,822 articles in 2018 to 103,766 in 2022, BRICS output increased
from 104,654 to 171,713 in the same five-year period. Furthermore, OECD
publication output in Health Science Journals and Life Science journals
stagnated across the investigated hybrid journal portfolios after a peak
in 2020.

To illustrate the situation in 2022, \ref{fig:country_patch}B compares
total publication output with the number of open access articles. With
391,530 articles, China was the most productive country, followed by the
United States (268,965 articles) and India (87,428 articles). In
contrast, West and Nord European countries published a considerable high
number of open access articles, mainly due to transformative agreements.
Particularly, Germany, Great Britain, the Netherlands, Sweden,
Switzerland and Spain recorded an above-average open access share as
indicated by the linear trend line. As represented by the point size, as
well as it can been seen in Figure \ref{fig:country_patch}C,
transformative agreements contributed to this market position of these
countries. Interestingly, the United States had a notable open access
market share of 15\%, although transformative agreements contributed to
a lesser extent. Similarly, China's open access market share of 7.2 in
2022 was comparable to that of the Netherlands, which was (7.1\%,).

{\centering \includegraphics[width=0.99\linewidth,]{fig/country_top_20_plot-1}

\caption{Top 20 countries}\label{fig:country_top_20_plot}
\end{figure}

Figure \ref{fig:country_top_20_plot} illustrates the development of
hybrid open access from 2018 to 2022, highlighting the top 20 most
productive countries in terms of articles published in hybrid journals
that were included in transformative agreements over the five-year
period. Notably, The Netherlands (27\%), Sweden (24\%), Poland (17\%)
and Great Britain (17\%)) exhibited a relatively high level of uptake in
2018 which continued to increase in the following years. In 2022, Sweden
had the highest proportion of open-access articles relative to its
publication output (78\%), followed by the Netherlands (67\%) and
Switzerland (57\%), with these countries benefiting from transformative
agreements. In Germany, however, hybrid open access only began to
increase from 2019 onwards after the successful negotiation of
nationwide agreements with Wiley (July 2019) and Springer Nature
(January 2020). Prior to this, only a few organisations had agreements
in place, for the example the Max Planck Society with Springer Compact.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Bergstrom_2014}{}}%
Bergstrom, T. C., Courant, P. N., McAfee, R. P., \& Williams, M. A.
(2014). Evaluating big deal journal bundles. \emph{Proceedings of the
National Academy of Sciences}, \emph{111}(26), 9425--9430.
\url{https://doi.org/10.1073/pnas.1403006111}

\leavevmode\vadjust pre{\hypertarget{ref-Bj_rk_2012}{}}%
Björk, B.-C. (2012). The hybrid model for open access publication of
scholarly articles: A failed experiment? \emph{Journal of the American
Society for Information Science and Technology}, \emph{63}(8),
1496--1504. \url{https://doi.org/10.1002/asi.22709}

\leavevmode\vadjust pre{\hypertarget{ref-Bj_rk_2017}{}}%
Björk, B.-C. (2017). Growth of hybrid open access, 2009--2016.
\emph{PeerJ}, \emph{5}, e3878. \url{https://doi.org/10.7717/peerj.3878}

\leavevmode\vadjust pre{\hypertarget{ref-BJ_RK_2014}{}}%
BJÖRK, B., \& SOLOMON, D. (2014). How research funders can finance APCs
in full OA and hybrid journals. \emph{Learned Publishing}, \emph{27}(2),
93--103. \url{https://doi.org/10.1087/20140203}

\leavevmode\vadjust pre{\hypertarget{ref-Borrego_2020}{}}%
Borrego, Á., Anglada, L., \& Abadal, E. (2020). Transformative
agreements: Do they pave the way to open access? \emph{Learned
Publishing}. \url{https://doi.org/10.1002/leap.1347}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Dallmeier_2010}{}}%
Dallmeier-Tiessen, S., Goerner, B., Darby, R., Hyppoelae, J.,
Igo-Kemenes, P., Kahn, D., Lambert, S., Lengenfelder, A., Leonard, C.,
Mele, S., Polydoratou, P., Ross, D., Ruiz-Perez, S., Schimmer, R.,
Swaisland, M., \& Stelt, W. van der. (2010). \emph{{Open Access
Publishing - Models and Attributes}} (T. S. consortium, Ed.).

\leavevmode\vadjust pre{\hypertarget{ref-Fraser_2023}{}}%
Fraser, N., Hobert, A., Jahn, N., Mayr, P., \& Peters, I. (2023). No
deal: German researchers' publishing and citing behaviors after big deal
negotiations with elsevier. \emph{Quantitative Science Studies},
\emph{4}(2), 325--352. \url{https://doi.org/10.1162/qss_a_00255}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Hinchliffe_2019}{}}%
Hinchliffe, L. J. (2019). \emph{Transformative agreements: A primer}.
\url{https://web.archive.org/web/20210128170342/https://scholarlykitchen.sspnet.org/2019/04/23/transformative-agreements/};
The Scholarly Kitchen.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Jahn_2016}{}}%
Jahn, N., \& Tullney, M. (2016). A study of institutional spending on
open access publication fees in germany. \emph{{PeerJ}}, \emph{4},
e2323. \url{https://doi.org/10.7717/peerj.2323}

\leavevmode\vadjust pre{\hypertarget{ref-Jubb_2017}{}}%
Jubb, M., Plume, A., Oeben, S., Brammer, L., Johnson, R., Bütün, C., \&
Pinfield, S. (2017). \emph{Monitoring the transition to open access:
December 2017}.
\url{https://web.archive.org/web/20200212015524/https://www.universitiesuk.ac.uk/policy-and-analysis/reports/Documents/2017/monitoring-transition-open-access-2017.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-Laakso_2016}{}}%
Laakso, M., \& Björk, B.-C. (2016). Hybrid open access--a longitudinal
study. \emph{Journal of Informetrics}, \emph{10}(4), 919--932.
\url{https://doi.org/10.1016/j.joi.2016.08.002}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Larivi_re_2015}{}}%
Larivière, V., Haustein, S., \& Mongeon, P. (2015). The oligopoly of
academic publishers in the digital era. \emph{{PLOS} {ONE}},
\emph{10}(6), e0127502.
\url{https://doi.org/10.1371/journal.pone.0127502}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Pinfield_2016}{}}%
Pinfield, S., Salter, J., \& Bath, P. A. (2016). The "total cost of
publication" in a hybrid open-access environment: Institutional
approaches to funding journal article-processing charges in combination
with subscriptions. \emph{Journal of the Association for Information
Science and Technology}, \emph{67}(7), 1751--1766.
\url{https://doi.org/10.1002/asi.23446}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Robinson_Garcia_2020}{}}%
Robinson-Garcia, N., Costas, R., \& Leeuwen, T. N. van. (2020). Open
access uptake by universities worldwide. \emph{{PeerJ}}, \emph{8},
e9410. \url{https://doi.org/10.7717/peerj.9410}

\leavevmode\vadjust pre{\hypertarget{ref-Schiltz_2018}{}}%
Schiltz, M. (2018). Science without publication paywalls: cOAlition s
for the realisation of full and immediate open access. \emph{PLOS
Biology}, \emph{16}(9), e3000031.
\url{https://doi.org/10.1371/journal.pbio.3000031}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Shu_2023}{}}%
Shu, F., \& Larivière, V. (2023). The oligopoly of open access
publishing. \emph{Scientometrics}, \emph{129}(1), 519--536.
\url{https://doi.org/10.1007/s11192-023-04876-2}
